Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf forest

### 2019-11-17 manuscript/climate_sensitivity_draft.Rmd
Consolidate author feedback to end of methods
#65

NOT including Lawren's yet

- Individual-level drought resistance decreased with tree height, whereas crown position and root water access explained little variation. Resistance was lower among species whose leaves had a higher percent loss of area upon dehydration and lost turgor at less negative water potential. However, there was substantial variation in the best predictor variables across the three drought periods. **I think Alan's comments here warrant discussion. I believe the answer to his second comment is yes. In addition, Erika suggests spelling out the best predictor traits despite the point below. Thoughts?**

- We conclude that height--as opposed to canopy position, root water access, or species' traits--is the primary factor through which tree size mediates drought response, that leaf hydraulic traits are useful indicators of drought sensitivity, and that directions but not strengths of responses are consistent across droughts.

*Key words*:  canopy position; drought; Forest Global Earth Observatory (ForestGEO); leaf hydraulic traits; temperate broadleaf deciduous forest; tree growth; tree height; tree-ring [5-8] *Five to eight key words (in alphabetical order) . Words that are in the title can, and should, be among these.* **Erika is of the opinion that broadleaf should not be included here, instead using "deciduous". She also added deciduous to the title. I'm unsure if this is redundant**

\newpage
### Introduction
**Nobby says this intro is based too much on Bennet 2015. I'm not sure if he means we should remove some of the citations or rephrase some sentences.**
Forests globally play a critical role in climate regulation [@bonan_forests_2008], yet there remains enormous uncertainty as to how the terrestrial carbon (C) sink, which is dominated by forests, will respond to climate change [@friedlingstein_climatecarbon_2006]. An important aspect of this uncertainty lies with physiological responses of trees to drought [@kennedy_implementing_2019]. In many forested regions around the world, the risk of severe drought is increasing [@trenberth_global_2014; @cook_global_2014; @cook_unprecedented_2015; @mankin_mid-latitude_2019], sometimes despite increasing precipitation [@intergovernmental_panel_on_climate_change_climate_2015]. Droughts resulting from global change have been affecting forests worldwide [@allen_global_2010], and are expected to continue severely impacting forests [@anderegg_consequences_2013; @adams_temperature_2009; @breda_temperate_2006; @hartmann_research_2018; @allen_underestimation_2015; @abrams_interdisciplinary_2016; @adams_multi-species_2017]. During drought, larger trees generally suffer greater reductions in growth and increased mortality rates [e.g., @bennett_larger_2015; @stovall_tree_2019], resulting in disproportionate impacts on forest C storage [@meakem_role_2018]. As a result, forest drought responses stand to strongly impact forest feedbacks to climate change [@mcdowell_multi-scale_2016; @charney_observed_2016]. Yet, accurate characterization of drought responses remains a modeling challenge [@kennedy_implementing_2019]-- in part because some of the underlying mechanisms remain unclear (**REF**). Understanding forest responses to drought requires increased functional understanding of how tree size, microhabitat, and species' traits jointly confer individual-level vulnerability or resistance, and the extent to which their influence is consistent across droughts.

One fundamental question regarding forest responses to drought is what drives the observed tendency for large trees to suffer more during drought. @bennett_larger_2015 was the first to show larger growth reductions for larger trees on a global scale, and numerous subsequent studies have reinforced this finding (e.g., @stovall_tree_2019; @hacket-pain_consistent_2016]. However, they quantified tree size based on DBH, which has no direct mechanistic response. Their study proposed two major mechanisms--besides the tendency for bark beetles to preferentially attack larger trees [@pfeifer_observations_2011]--for the observed greater drought growth reductions of large trees **Alan gives another mechanism here from another paper; I'm unsure how you want to consolidate that with this sentence.**. First, @bennett_larger_2015 found taller trees have a greater biophysical challenge of lifting water greater distances against the effects of gravity and friction [@mcdowell_relationships_2011; @mcdowell_darcys_2015; @ryan_hydraulic_2006; @couvreur_water_2018], especially during drought [@zhang_size-dependent_2009]. Second, larger trees may have lower drought resistance because of their higher crown positions, where they are exposed to higher solar radiation, greater wind speeds, lower humidity, and lower CO_2_ concentrations [@koike_leaf_2001; @kunert_revised_2017 **REFS-KAT**). Subcanopy trees subsequently tend to fare better specifically due to the benefits of a buffered environment [@pretzsch_drought_2018]. Potentially counteracting the biophysical challenges faced by large trees, their larger root systems confer an advantage in terms of allowing greater access to water; however, it appears that this effect is usually insufficient to offset the costs of height and/or crown exposure [@bennett_larger_2015]. **For this previous sentence, Alan questions whether this assumption can be made, and Valentine says it's odd to come back to Bennett as a ref here.** In addition to what was described by @bennett_larger_2015, a final mechanism that could mediate tree size-related responses to drought is how hydraulic traits are distributed with respect to size [@meakem_role_2018; @liu_hydraulic_2019]. The pattern observed by @bennett_larger_2015 could be caused if the larger size classes were dominated by species less adapted to handle drought, be it through avoidance, resistance, or resilience **Valentine is confused if you mean to say "lack of" these things. Alan points out we should define these terms here since is the first time they're used. Unsure what your thoughts are on that**. Alternatively, larger size classes may be dominated by species that are inherently better adapted to greater biophysical challenges--as is the case in tropical moist forests of Panama, where larger size classes contain greater proportions of deciduous species [@condit_quantifying_2000; @meakem_role_2018]. Understanding the mechanisms driving more severe suffering of larger trees during drought will require sorting out the interactive effects of height, canopy position, root water access, and species' traits.

@@

A final fundamental question regarding forest responses to drought is whether tree size and species' traits have similar influence across droughts, or whether variability in factors such as drought severity, duration, and timing interact with tree size and traits such that components of the community respond differently across droughts. No two droughts are the same, and tree growth responses vary with drought characteristics such as timing and atmospheric demand [@dorangeville_drought_2018]. However, we are not aware of any studies that compare how tree size and species' traits mediate growth responses across droughts. While tree-ring studies provide long-term records of tree responses to multiple droughts (e.g., [@lloret_components_2011;  @dorangeville_drought_2018] **REFS**), these don't test for differential trait effects across periods of water shortage [@dorangeville_drought_2018] and generally focus on species-level responses, which preclude consideration of the roles of tree size and microenvironment. The ecological studies that have shaped our understanding of the role of tree size and microenvironment in forest drought responses generally examine only a single drought and tend to focus disproportionately on extreme droughts with dramatic impacts (*e.g.*, [@allen_global_2010; @bennett_larger_2015; @stovall_tree_2019; @anderegg_meta-analysis_2016). Thus, our knowledge of forest responses to more modest but frequent droughts--e.g., those with historical return intervals on the order of a decade--remains more limited. While the tendency for larger trees to suffer more certainly predominates [@bennett_larger_2015], there are exceptions (*e.g.*, **REFS??**). There is also evidence that the degree to which larger trees suffer more increases with the severity of drought conditions [@bennett_larger_2015; @stovall_tree_2019]. [*Are there any studies showing interactions of drought type with traits?*] Thus, while we expect many of the fundamental mechanisms shaping drought responses to be universal, we have little undertanding of how tree size and traits interact with drought characteristics to result in differential responses across droughts.

Here, we combine tree-ring records covering three droughts (1966, 1977, 1999), species functional and hydraulic trait measurements, and forest census data from a 25.6-ha ForestGEO plot in Virginia (USA) to test a series of hypotheses and associated specific predictions (Table 1) designed to yield functional understanding of how tree size, microenvironment, and species' traits collectively shape drought responses. 
First, we focus on the role of tree size and its interaction with microenvironment. 
We confirm that, consistent with most forests globally, larger-diameter trees tend to have lower drought resistance in this forest, which is in an ecoregion represented by only one study in [@bennett_larger_2015] (*H1.0*). 
We then test hypotheses designed to disentangle the relative importance of tree height (*H1.1*), crown exposure (*H1.2*), and root water access, which should be greater for larger trees in dry but not in perpetually wet microsites (*H1.3*). 
Second, we focus on the role of species' functional and hydraulic traits and their interaction with tree height. We hypothesize that drought resistance will follow observed patterns in relation to wood density (negative effect; @hoffmann_hydraulic_2011), specific leaf area (positive effect), and xylem architecture (*i.e.*, ring or diffuse/ semi-ring porous), but that hydraulic leaf traits including $PLA_{dry}$ and $\pi_{tlp}$ will prove better predictors (*H2.1*). 
We then test whether these traits correlate with tree height (*H2.2*), potentially driving the observed tendency for taller trees to suffer more during drought (*H2.3*). 
Finally, we focused on variability among droughts, asking how community resistance varied across droughts (*H3.1*) and whether the factors confirming vulnerability or resistance varied across droughts (*H3.2*).

@@

Research was conducted at the 25.6 ha ForestGEO (Forest Global Earth Observatory) study plot at the Smithsonian Conservation Biology Institute (SCBI) in Virginia, USA (38\°53'36.6"N, 78\°08'43.4"W) [@bourg_initial_2013; @andersonteixeira_ctfs-forestgeo:_2015]. SCBI is located in the central Appalachian Mountains at the northern edge of Shenandoah National Park.  Elevations range from 273-338m above sea level [@gonzalezakre_patterns_2016] with a topographic relief of 65m [@bourg_initial_2013]. Climate is humid temperate, with mean annual temperature of 12.7^$\circ$^C and precipitation of 1005 mm during our study period (1960-2009; source: CRU TS v.4.01; @harris_updated_2014). Dominant tree taxa include *Liriodendron tulipifera*,  oaks (*Quercus* spp.), and hickories (*Carya* spp.).

@@

Within or just outside the ForestGEO plot, we collected data on a suite of variables including tree size, microenvironemnt, and species traits (Table 2). The SCBI ForestGEO plot was censused in 2008, 2013, and 2018 following standard ForestGEO protocols, whereby all free-standing woody stems $\ge$ 1cm diameter at breast height (DBH) were mapped, tagged, measured at DBH, and identified to species [@condit_tropical_1998]. From this census data, we used measurements of DBH from 2008 to calculate historical DBH, tree location in the plot to determine the topographic wetness index, and data for all stems $\ge$ 10cm to analyze functional trait composition relative to tree height (all analyses described below). Census data, which were last updated in 2019, are available through the ForestGEO data portal (www.forestgeo.si.edu).

We analyzed tree-ring data from 571 trees representing the twelve species contributing most to woody aboveground net primary productivity ($ANPP_{stem}$), which together comprised 97% of study plot $ANPP_{stem}$ between 2008 and 2013 [@helcoski_growing_2019] (Figure S1). Cores were obtained in 2010-2011 or 2016-2017 from a breast height of 1.3m using a 5mm increment borer. In 2010-2011, cores were collected from randomly selected live trees of species with at least 30 individuals of DBH $\ge$ 10cm [@bourg_initial_2013]. In 2016-2017, cores were collected from all trees found dead in the annual mortality census [@gonzalezakre_patterns_2016]. Cores were sanded, measured, and cross-dated using standard procedures, as detailed in [@helcoski_growing_2019]. The resulting chronologies have been published in association with @helcoski_growing_2019: (ITRDB; GitHub/Zenodo).  **Ryan, have you heard back from ITRDB?  We should also cite GitHub/Zenodo here. **

For each tree, we combined tree-ring records and allometric equations of bark thickness to retroactively calculate DBH for the years 1950-2009. 
Prior DBH was estimated using the following equation:

@@

Here, $Y$ denotes the year of interest, $r_{ring}$ denotes ring width derived from cores, and $r_{bark}$ denotes bark thickness. 
Bark thickness was estimated from species-specific allometries based on the bark thickness data from the site [@andersonteixeira_size-related_2015].
Specifically, we used linear regression equations on log-transformed data to relate bark thickness to diameter without bark from 2008 data (Table S1), which were then used to determine bark thickness in the retroactive calculation of DBH.

Height measurements (n=1518 trees) were taken by several researchers between 2012 to 2019, and are archived in a public [GitHub repository](https://github.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/tree/master/tree_dimensions/tree_heights). Measurement methods included manual [@stovall_assessing_2018, NEON], digital rangefinders [@andersonteixeira_size-related_2015], and automatic LiDAR [@stovall_terrestrial_2018]. Rangefinders used either the tangent method (Impulse 200LR, TruPulse 360R) or the sine method (Nikon ForestryPro) for calculating heights. Both methods are associated with some error [@larjavaara_measuring_2013], but in this instance there was no clear advantage of one or the other. Species-specific height allometries were developed (Table S2). For species with insufficient height data to create reliable species-specific allometries, heights were calculated from an equation developed using all height measurements.

Crown positions were recorded in the field during the growing season of 2018 following the crown position protocol of @jennings_assessing_1999, whereby positions were ranked as dominant, codominant, intermediate, or suppressed. As there was no way to retroactively estimate crown position, we assumed that 2018 crown position was reflective of each tree's position over the past 60 years. While some trees undoubtedly changed position, an analysis of crown position relative to height (Fig. 2) and height change since the beginning of the study period indicated that change was likely slow (Fig. S3). Specifically, average tree height growth was confined to ~0.82m from 1966 to 1977, ~1.45m from 1977 to 1999, and ~1.97m from 1999 to 2018.

@@

Hydraulic traits were collected at SCBI (Table 3). In August 2018, we sampled small sun-exposed branches from three individuals of each species in and around the ForestGEO plot. Samples were re-cut under water by at least two notes and re-hydrated overnight in buckets covered with opaque plastic bags before measurements. Rehydrated leaves (n=3 per indivdiual) were scanned, weighed, dried at 60$^\circ$ C for $\ge$ 48 hours, and then re-scanned and weighed. Leaf area was calculated from scanned images using the LeafArea R package [@R-LeafArea]. $LMA$ was calculated as the ratio of leaf dry mass to fresh area. $PLA$ was calculated as the percent loss of area between fresh and dry leaves. $WD$ was calculated for ~1cm diameter stem samples (bark and pith removed) as the ratio of dry weight to volume. We used the rapid determination method of @bartlett_rapid_2012 to estimate water potential at turgor loss point ($\pi_{tlp}$). Briefly, two 4mm diameter leaf discs were cut from each leaf, tightly wrapped in foil, submerged in liquid nitrogen, perforated 10-15 times with a dissection needle, and then measured using a vapour pressure osmometer (VAPRO 5520, Wescor, Logan, UT, USA). Osmotic potential ($\pi_{osm}$) given by the osmometer was used to estimate ($\pi_{tlp}$) using the equation $\pi_{tlp}=0.832 \pi_{osm} ^{-0.631}$ [@bartlett_rapid_2012]. Data and R scripts for hydraulic traits are available at [**create new public GitHub repo for hydraulic traits, archive in Zenodo, give DOI**].

To characterize how environmental conditions vary with height, data were obtained from the National Ecological Observation Network (NEON) tower located <1km from the study area. We used data on wind speed, relative humidity, and air temperature, all measured over a vertical profile, for the years 2016-2018 [@noauthor_national_2018]. After filtering for missing and outlier values, the data was consolidated to represent the mean values per sensor height per day.

@@

We identified droughts within the time period 1950-2009, defining drought [@slette_how_2019] as events with widespread reductions in tree growth and anomalously dry peak growing season climatic conditions. Specifically, candidate drought years were defined as those where >25% of the cored trees experienced >30% reduction in basal area increment (BAI) relative to the previous 5 years, following the drought resistance ($R$) metric of [@lloret_components_2011], and identified using the pointRes package [@R-pointRes] in R (version 3.5.3). 
Separately, we identified the years with driest conditions during May-August, which stood out in the analysis of [@helcoski_growing_2019] as the current-year months to which annual growth was most sensitive for trees at this site. We considered two metrics of moisture deficit: NOAA Divisional Data’s Palmer Drought Severity Index (PDSI) and the difference between monthly potential evapotranspiration (PET) and precipitation (PRE). These data were obtained from the ForestGEO Climate Data Portal (https://github.com/forestgeo/Climate) in August 2018, with monthly PET and PRE sourced from Climatic Research Unit high-resolution gridded dataset (CRU TS v.4.01; @harris_updated_2014). The driest years were identified through simply ranking mean May-August PDSI or [PET-PRE] for the time period from driest to wettest.

*Analysis*

For each drought period, we calculated drought resistance ($R$) as the ratio of BAI during drought to that of the five previous years [@lloret_components_2011]. Analyses focused on testing the predictions presented in Table 1, most of which consider $R$ as the response variable. The general statistical model for hypothesis testing was a mixed effects model (lme4 package from @R-lme4) with $R$ as the response variable, tree nested within species as a random effect, and one or more independent variables as fixed effects. We used AICc (AICcmodavg package from @R-AICcmodavg) to assess model fit. Models were run for all drought years combined (with year as a fixed effect) and for each drought year independently. We first tested the predictor variables independently, counting a variable as useful for predicting $R$ when AICc was reduced by $\ge$ 2 relative to the appropriate null model lacking that variable (Table 4). Because height had a substantial influence on $R$, we included it in the null model for testing other variables (Table 4).

Second, for each drought year and for all years combined, we determined the best multivariate models for predicting $R$. All variables with dAICc > 1.0 in any model (drought years combined or any individual drought) were used as candidate variables for each scenario's best model. We compared models with all possible combinations of candidate variables and identified the full set of models within dAICc=2 of the top model (that with lowest AICc). These were counted as the "top multivariate models". When a variable appeared in all of these top models and the sign of the coefficient was consistent across models, this was counted as support for/ rejection of the associated prediction by the multivariate models. If the variable appeared in only some of the models, we considered this partial support.

>>>>

- Individual-level drought resistance decreased with tree height, whereas crown position and soil water availability explained little variation. Resistance was lower among species whose leaves had a higher percent loss of area upon dehydration and lost turgor at less negative water potential. However, there was substantial variation in the best predictor variables across the three drought periods. **See comment 1 and 2**

- We conclude that height--as opposed to canopy position, soil water availability, or species' traits--is the primary factor through which tree size mediates drought response, that leaf hydraulic traits are useful indicators of drought sensitivity, and that directions but not strengths of responses are consistent across droughts.

*Key words*:  canopy position; drought; Forest Global Earth Observatory (ForestGEO); leaf hydraulic traits; temperate broadleaf deciduous forest; tree growth; tree height; tree-ring [5-8] *Five to eight key words (in alphabetical order) . Words that are in the title can, and should, be among these.* **See comment 3 **

\newpage
### Introduction
**See comment 4**
Forests globally play a critical role in climate regulation [@bonan_forests_2008], yet there remains enormous uncertainty as to how the terrestrial carbon (C) sink, which is dominated by forests, will respond to climate change [@friedlingstein_climatecarbon_2006]. An important aspect of this uncertainty lies with physiological responses of trees to drought [@kennedy_implementing_2019]. In many forested regions around the world, the risk of severe drought is increasing [@trenberth_global_2014; @cook_global_2014; @cook_unprecedented_2015; @mankin_mid-latitude_2019], sometimes despite increasing precipitation [@intergovernmental_panel_on_climate_change_climate_2015]. Droughts resulting from global change have been affecting forests worldwide [@allen_global_2010], and are expected to continue severely impacting forests [@anderegg_consequences_2013; @adams_temperature_2009; @breda_temperate_2006; @hartmann_research_2018; @allen_underestimation_2015; @abrams_interdisciplinary_2016; @adams_multi-species_2017]. During drought, larger trees generally suffer greater reductions in growth and increased mortality rates [e.g., @bennett_larger_2015; @stovall_tree_2019], resulting in disproportionate impacts on forest C storage [@meakem_role_2018]. As a result, forest drought responses stand to strongly impact forest feedbacks to climate change [@mcdowell_multi-scale_2016; @charney_observed_2016]. Yet, accurate characterization of drought responses remains a modeling challenge [@kennedy_implementing_2019]-- in part because some of the underlying mechanisms remain unclear (**REF**). Understanding forest responses to drought requires increased functional understanding of how tree size, microhabitat, and species' traits jointly confer individual-level vulnerability or resistance, and the extent to which their influence is consistent across droughts.

One fundamental question regarding forest responses to drought is what drives the observed tendency for large trees to suffer more during drought. @bennett_larger_2015 was the first to show larger growth reductions for larger trees on a global scale, and numerous subsequent studies have reinforced this finding (e.g., @stovall_tree_2019; @hacket-pain_consistent_2016]. However, they quantified tree size based on DBH, which has no direct mechanistic response. Their study proposed two major mechanisms--besides the tendency for bark beetles to preferentially attack larger trees [@pfeifer_observations_2011]--for the observed greater drought growth reductions of large trees **see comment 5**. First, @bennett_larger_2015 found taller trees have a greater biophysical challenge of lifting water greater distances against the effects of gravity and friction [@mcdowell_relationships_2011; @mcdowell_darcys_2015; @ryan_hydraulic_2006; @couvreur_water_2018], especially during drought [@zhang_size-dependent_2009]. Second, larger trees may have lower drought resistance because of their higher crown positions, where they are exposed to higher solar radiation, greater wind speeds, lower humidity, and lower CO_2_ concentrations [@koike_leaf_2001; @kunert_revised_2017 **REFS-KAT**). Subcanopy trees subsequently tend to fare better specifically due to the benefits of a buffered environment [@pretzsch_drought_2018]. Potentially counteracting the biophysical challenges faced by large trees, their larger root systems confer an advantage in terms of allowing greater access to water; however, it appears that this effect is usually insufficient to offset the costs of height and/or crown exposure [@bennett_larger_2015]. **see comment 6** In addition to what was described by @bennett_larger_2015, a final mechanism that could mediate tree size-related responses to drought is how hydraulic traits are distributed with respect to size [@meakem_role_2018; @liu_hydraulic_2019]. The pattern observed by @bennett_larger_2015 could be caused if the larger size classes were dominated by species less adapted to handle drought, be it through avoidance, resistance, or resilience **see comment 7**. Alternatively, larger size classes may be dominated by species that are inherently better adapted to greater biophysical challenges--as is the case in tropical moist forests of Panama, where larger size classes contain greater proportions of deciduous species [@condit_quantifying_2000; @meakem_role_2018]. Understanding the mechanisms driving more severe suffering of larger trees during drought will require sorting out the interactive effects of height, canopy position, soil water availability, and species' traits.

@@

A final central question regarding forest responses to drought is whether tree size and species' traits have similar influence across different drought events, or whether that influence is more strongly predicted by community-level responses to variable drought severity, duration, and timing based on tree size and traits. Tree growth responses vary with drought characteristics such as timing and atmospheric demand [@dorangeville_drought_2018], but the question of how tree size and species' traits impact growth responses across droughts still remains. While tree-ring studies provide long-term records of tree responses to multiple droughts (e.g., [@lloret_components_2011;  @dorangeville_drought_2018] **REFS**), they *don't test for differential trait effects across periods of water shortage [@dorangeville_drought_2018] and generally* **see comment 8** focus on species-level responses, and do not consider the roles of tree size and microenvironment. The ecological field-baesd studies that have shaped our understanding of the role of tree size and microenvironment in forest drought responses generally examine only a single drought and tend to focus disproportionately on extreme droughts with dramatic impacts (*e.g.*, [@allen_global_2010; @bennett_larger_2015; @stovall_tree_2019; @anderegg_meta-analysis_2016). Thus, our knowledge of forest responses to more modest but frequent droughts - e.g., those with historical return intervals on the order of one to two decades - remains limited. While the tendency for larger trees to experience more severe drought certainly predominates the literature [@bennett_larger_2015], there are exceptions [*e.g.*, @suarez_factors_2004; @zang_size-dependent_2012; @mcdowell_mechanisms_2008]. There is also evidence that the degree to which larger trees are more impacted by drought increases with the severity of drought conditions [@bennett_larger_2015; @stovall_tree_2019]. [*Are there any studies showing interactions of drought type with traits?*] **see comment 9** Thus, while we expect many of the mechanisms shaping drought responses to be universal, we have little understanding of how tree size and traits interact with drought characteristics.

To yield functional understanding of how tree size, microenvironment, and species' traits collectively shape drought responses, we test a series of hypotheses and associated specific predictions (Table 1) based on the combination of tree-ring records from three droughts (1966, 1977, 1999), species functional and hydraulic trait measurements, and forest census data from a 25.6-ha ForestGEO plot in Virginia (USA). First, we focus on the role of tree size and its interaction with microenvironment. We confirm that, consistent with most forests globally, larger-diameter trees tend to have lower drought resistance in this forest, which is in an ecoregion **see comment 10** represented by only one study in [@bennett_larger_2015] (*H1.0*). 
We then test hypotheses designed to disentangle the relative importance of tree height (*H1.1*), crown exposure (*H1.2*), and soil water availability, which should be greater for larger trees in dry but not in perpetually wet microsites (*H1.3*). **see comments 11,12**. 
Second, we focus on the role of species' functional and hydraulic traits and their interaction with tree height. We hypothesize that drought resistance will be negatively affected by wood density (following observed patterns by **e.g.** @hoffmann_hydraulic_2011), and positively affected by specific leaf area and xylem architecture (*i.e.*, ring or diffuse/ semi-ring porous), but that hydraulic leaf traits including $PLA_{dry}$ and $\pi_{tlp}$ will prove better predictors (*H2.1*). 
We then test whether these traits correlate with tree height (*H2.2*), potentially driving the observed tendency for taller trees to suffer more during drought after accounting for trait effects (*H2.3*). 
Finally, we focused on variability among droughts, asking how community resistance varied across droughts (*H3.1*) and whether the factors confirming vulnerability or resistance varied across droughts (*H3.2*).

@@

Research was conducted at the 25.6 ha ForestGEO (Forest Global Earth Observatory) study plot at the Smithsonian Conservation Biology Institute (SCBI) in Virginia, USA (38\°53'36.6"N, 78\°08'43.4"W) [@bourg_initial_2013; @andersonteixeira_ctfs-forestgeo:_2015]. SCBI is located in the central Appalachian Mountains at the northern edge of Shenandoah National Park.  Elevations range from 273-338m above sea level with a topographic relief of 65m [@bourg_initial_2013]. Climate is humid temperate, with mean annual temperature of 12.7^$\circ$^C and precipitation of 1005 mm during our study period (1960-2009; source: CRU TS v.4.01; @harris_updated_2014). Dominant tree taxa within this secondary forest include *Liriodendron tulipifera*,  oaks (*Quercus* spp.), and hickories (*Carya* spp.). **see comment 13**

@@

All analysis beyond basic data collection was performed using R version 3.5.3 [@R-base].

Within or just outside the ForestGEO plot, we collected data on a suite of variables including tree size, microenvironment, and species traits (Table 2). Destructive sampling was only carried out within the plot. **see comment 14** The SCBI ForestGEO plot was censused in 2008, 2013, and 2018 following standard ForestGEO protocols, whereby all free-standing woody stems $\ge$ 1cm diameter at breast height (DBH) were mapped, tagged, measured at DBH, and identified to species [@condit_tropical_1998]. From this census data, we used measurements of DBH from 2008 to calculate historical DBH and data for all stems $\ge$ 10cm to analyze functional trait composition relative to tree height (all analyses described below). Census data, which were last updated in 2019, are available through the ForestGEO data portal (www.forestgeo.si.edu).

We analyzed tree-ring data (cambial growth increment) **see comment 15** from 571 trees representing the twelve species contributing most to woody aboveground net primary productivity ($ANPP_{stem}$), which together comprised 97% of study plot $ANPP_{stem}$ between 2008 and 2013 [@helcoski_growing_2019] (Figure S1). Cores were collected at breast heigh (1.3m) in 2010-2011 or 2016-2017 using a 5mm increment borer. In 2010-2011, cores were collected from randomly selected live trees of species with at least 30 individuals $\ge$ 10cm DBH [@bourg_initial_2013]. In 2016-2017, cores were collected from all trees found dead in the annual mortality census [@gonzalezakre_patterns_2016]. Cores were sanded, measured, and cross-dated using standard procedures, as detailed in [@helcoski_growing_2019]. The resulting chronologies have been published in association with @helcoski_growing_2019: (ITRDB; GitHub/Zenodo).  **Ryan, have you heard back from ITRDB?  We should also cite GitHub/Zenodo here. **

For each cored tree, we combined tree-ring records and allometric equations of bark thickness to retroactively calculate DBH for the years 1950-2009. 
Prior DBH was estimated using the following equation:

@@

Here, $Y$ denotes the year of interest, $r_{ring}$ denotes ring width derived from cores, and $r_{bark}$ denotes bark thickness. 
Bark thickness was estimated from species-specific allometries based on the bark thickness data from the site [@andersonteixeira_size-related_2015].
Specifically, we used linear regression equations on log-transformed data to relate bark thickness to diameter inside bark from 2008 data (Table S1), which were then used to determine bark thickness in the retroactive calculation of DBH.

Height measurements (n=1518 trees) were taken by several researchers between 2012 to 2019, and are archived in a public [GitHub repository](https://github.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/tree/master/tree_dimensions/tree_heights). Measurement methods included manual [@stovall_assessing_2018, NEON], digital rangefinders [@andersonteixeira_size-related_2015], and automatic, ground-based LiDAR [@stovall_terrestrial_2018]. Rangefinders used either the tangent method (Impulse 200LR, TruPulse 360R) or the sine method (Nikon ForestryPro) for calculating heights. Both methods are associated with some error [@larjavaara_measuring_2013], but in this instance there was no clear advantage of one or the other. Species-specific height allometries were developed (Table S2). For species with insufficient height data to create reliable species-specific allometries, heights were calculated from an equation developed using all height measurements. **see comments 16,17**

Crown positions were recorded in the field during the growing season of 2018 following the crown position protocol of @jennings_assessing_1999, whereby positions were ranked as dominant, codominant, intermediate, or suppressed. As it was unfeasible to retroactively estimate crown position, we assumed that 2018 crown position was reflective of each tree's position over the past 60 years. While some trees undoubtedly changed position, an analysis of crown position relative to height (Fig. 2d) and height variance since the beginning of the study period indicated that change was likely slow (Fig. S3). **see comment 18,19** Specifically, average tree height growth was confined to ~0.82m from 1966 to 1977, ~1.45m from 1977 to 1999, and ~1.97m from 1999 to 2018.

@@

Hydraulic traits were collected at SCBI (Table 3) in August 2018. We sampled small sun-exposed branches **see comment 20** from three individuals of each species in and around the ForestGEO plot. Sampled branches were re-cut under water at least two nodes above the original cut and re-hydrated overnight in covered buckets (opaque plastic bags) before measurements were taken. Rehydrated leaves (n=3 per indivdiual) were scanned, weighed, dried at 60$^\circ$ C for $\ge$ 48 hours, and then re-scanned and weighed. Leaf area was calculated from scanned images using the LeafArea R package [@R-LeafArea]. $LMA$ was calculated as the ratio of leaf dry mass to fresh area. $PLA$ was calculated as the percent loss of area between fresh and dry leaves. $WD$ was calculated for ~1cm diameter stem samples (bark and pith removed) as the ratio of dry weight to volume. **see comment 21** We used the rapid determination method of @bartlett_rapid_2012 to estimate water potential at turgor loss point ($\pi_{tlp}$). Briefly, two 4mm diameter leaf discs were cut from each leaf, tightly wrapped in foil, submerged in liquid nitrogen, perforated 10-15 times with a dissection needle, and then measured using a vapour pressure osmometer (VAPRO 5520, Wescor, Logan, UT, USA). Osmotic potential ($\pi_{osm}$) given by the osmometer was used to estimate ($\pi_{tlp}$) using the equation $\pi_{tlp}=0.832 \pi_{osm} ^{-0.631}$ [@bartlett_rapid_2012]. Data and R scripts for hydraulic traits are available at [**create new public GitHub repo for hydraulic traits, archive in Zenodo, give DOI**]. **see comment 22**

To characterize how environmental conditions vary with height, data were obtained from the National Ecological Observation Network (NEON) tower located <1km from the study area. We used data on wind speed, relative humidity, and air temperature, all measured over a vertical profile **see comment 23**, for the years 2016-2018 [@noauthor_national_2018]. After filtering for missing and outlier values, the data was consolidated to represent the mean values per sensor height per day. The range of these means were then aggregated at a month scale.

@@

We identified droughts within the time period 1950-2009, defining drought [@slette_how_2019] as events with widespread reductions in tree growth and anomalously dry peak growing season climatic conditions. Specifically, candidate drought years were defined as those where >25% of the cored trees experienced >30% reduction in basal area increment (BAI) relative to the previous 5 years, following the drought resistance ($R$) metric of [@lloret_components_2011], and identified using the pointRes package [@R-pointRes] in R. **see comment 24**
Separately, we identified the years with driest conditions during May-August, which stood out in the analysis of [@helcoski_growing_2019] as the current-year months to which annual growth was most sensitive for trees at this site. We considered two metrics of moisture deficit: NOAA Divisional Data’s Palmer Drought Severity Index (PDSI) and the difference between monthly potential evapotranspiration (PET) and precipitation (PRE). These data were obtained from the ForestGEO Climate Data Portal (https://github.com/forestgeo/Climate) in August 2018, with monthly PET and PRE sourced from Climatic Research Unit high-resolution gridded dataset (CRU TS v.4.01; @harris_updated_2014). The driest years were identified through ranking mean May-August PDSI or [PET-PRE] for the time period from driest to wettest. **see comment 25**

*Statistical Analysis*

For each drought period, we calculated drought resistance ($R$) as the ratio of BAI during drought to the mean BAI over the five years preceding the drought [@lloret_components_2011]. Analyses focused on testing the predictions presented in Table 1, with either $R$ or $H$ as the response variable. **see comment 26** The general statistical model for hypothesis testing was a mixed effects model (lme4 package from @R-lme4) with $R$ as the response variable, tree nested within species as a random effect, and one or more independent variables as fixed effects. We used AICc (AICcmodavg package from @R-AICcmodavg) to assess model selection, and conditional/marginal R-squared to assess model fit.

Models were run for all drought years combined (with year as a fixed effect) and for each drought year independently. In order to determine the relative importance of the traits alone (H2.1), we first tested the predictor variables independently against both height and $R$ given height's substantial influence. Variables were considered useful for predicting $R$ when AICc was reduced by $\ge$ 2 relative to the appropriate null model lacking that variable (Table 4).

Once these beset predictor variables were determined, we calculated the best overall models for predicting $R$ across all traits, each individual drought year, and all years combined. We compared models with all possible combinations of candidate variables and identified the full set of models within dAICc=2 of the top model (that with lowest AICc). These were counted as the "top multivariate models". When a variable appeared in all of these top models and the sign of the coefficient was consistent across models, this was counted as support for/ rejection of the associated prediction by the multivariate models. If the variable appeared in only some of the models, we considered this partial support. **see comment 27**
